\bf Causal Inference in Population Trends: Searching for Demographic Anomalies in Big Data

### 2020-09-01 MainDocument/MainDocument.Rmd
Minor Edits.

abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in social data, previously dismissed as noise. Here, we combine causal inference techniques and abductive reasoning to identify fertility and mortality anomalies on twenty years of demographic data in the United States. We uncover real, “hidden” baby booms/busts and mortality spikes/dips, distinguishable from regular trend variations.  We identify more than 22 and 156 fertility and mortality anomalies, totaling more than 200k and 600k anomalous births and deaths, respectively. Notable detectable mortality anomalies include the September 11 2001 terrorist attack in New York, the emergence and acceleration of the opioid epidemic in New Hampshire, and the excess mortality due to Hurricane Maria in Puerto Rico. Notable fertility anomalies include the “missing births” in Louisiana after Hurricane Katrina and the reduction in fertility behavior after the September 2008 stock market crash in Connecticut, amongst others. Our combined causal inference and abductive reasoning approach can be readily adapted to find other, undiscovered social phenomena or to evaluate the efficacy of important public policies."

@@

>>>>

abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in social data, previously dismissed as noise. Here, we combine causal inference techniques and abductive reasoning to identify fertility and mortality anomalies on twenty years of complete demographic data in the United States. We uncover real, “hidden” baby booms/busts and mortality spikes/dips, distinguishable from regular trend variations. We identify more than 22 and 156 fertility and mortality anomalies, totaling more than 200k and 600k anomalous births and deaths, respectively. Notable detectable mortality anomalies include the September 11 2001 terrorist attack in New York, the emergence and acceleration of the opioid epidemic in New Hampshire, and the excess mortality due to Hurricane Maria in Puerto Rico. Notable fertility anomalies include the “missing births” in Louisiana after Hurricane Katrina and the reduction in fertility behavior after the September 2008 stock market crash in Connecticut, amongst others. Our combined causal inference and abductive reasoning approach can be readily adapted to find other, undiscovered social phenomena or to evaluate the efficacy of important public policies."

@@

## Replication Files
All data and code necessary to reproduce the reported results are licensed under the CC-BY-4.0 license and are publicly available in a replication repository located at https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1. The analyses were performed in *R* [@rcite].

### 2020-09-01 MainDocument/MainDocument.tex
Minor Edits.

\bigskip
\begin{abstract}
The proliferation of big data, wider access to advanced computing
platforms, and the development of powerful statistical algorithms can
uncover hidden anomalies in social data, previously dismissed as noise.
Here, we combine causal inference techniques and abductive reasoning to
identify fertility and mortality anomalies on twenty years of
demographic data in the United States. We uncover real, ``hidden'' baby
booms/busts and mortality spikes/dips, distinguishable from regular
trend variations. We identify more than 22 and 156 fertility and
mortality anomalies, totaling more than 200k and 600k anomalous births
and deaths, respectively. Notable detectable mortality anomalies include
the September 11 2001 terrorist attack in New York, the emergence and
acceleration of the opioid epidemic in New Hampshire, and the excess
mortality due to Hurricane Maria in Puerto Rico. Notable fertility
anomalies include the ``missing births'' in Louisiana after Hurricane
Katrina and the reduction in fertility behavior after the September 2008
stock market crash in Connecticut, amongst others. Our combined causal
inference and abductive reasoning approach can be readily adapted to
find other, undiscovered social phenomena or to evaluate the efficacy of
important public policies.
\end{abstract}

@@

>>>>

\bigskip
\begin{abstract}
The proliferation of big data, wider access to advanced computing
platforms, and the development of powerful statistical algorithms can
uncover hidden anomalies in social data, previously dismissed as noise.
Here, we combine causal inference techniques and abductive reasoning to
identify fertility and mortality anomalies on twenty years of complete
demographic data in the United States. We uncover real, ``hidden'' baby
booms/busts and mortality spikes/dips, distinguishable from regular
trend variations. We identify more than 22 and 156 fertility and
mortality anomalies, totaling more than 200k and 600k anomalous births
and deaths, respectively. Notable detectable mortality anomalies include
the September 11 2001 terrorist attack in New York, the emergence and
acceleration of the opioid epidemic in New Hampshire, and the excess
mortality due to Hurricane Maria in Puerto Rico. Notable fertility
anomalies include the ``missing births'' in Louisiana after Hurricane
Katrina and the reduction in fertility behavior after the September 2008
stock market crash in Connecticut, amongst others. Our combined causal
inference and abductive reasoning approach can be readily adapted to
find other, undiscovered social phenomena or to evaluate the efficacy of
important public policies.
\end{abstract}

@@

\hypertarget{replication-files}{%
\subsection{Replication Files}\label{replication-files}}

All data and code necessary to reproduce the reported results are
licensed under the CC-BY-4.0 license and are publicly available in a
replication repository located at
\url{https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1}.
The analyses were performed in \emph{R} \citep{rcite}.
